Is high social class always beneficial for survival? Long-term perspectives of social inequality in mortality among adults and elderly: Northern Sweden 1851--2013

### 2017-07-24 article.Rmd
Deleting sections income and education

In the present paper, we take a long-term perspective on the development of
social inequalities in mortality in the adult and elderly population. The
areas under investigation are the Skellefteå and Umeå regions in the north
of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013
for Umeå. The main question we investigate is if inequality in mortality
between social classes increased among the adult and elderly population,
primarily benefiting the highest social positions and how this is connected
to the development of economic inequality in society. This issue is then
analysed according to gender, age-group (working age vs retired) and cause
of death (cardiovascular and cancer). Furthermore, we compare our results
witn analyses of alternative inequality indicators, education and income
for the period 1986--2013.

@@

## Income

Income is based on disposable individual income from the LISA database at
Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to
2005.

## Education

Education is here analysed only for the period 1990 to 2013. Years of
education are given in 3 categories and is based on level of education
according to categories in the LISA database at Statistics Sweden. It
ranges from basic education (nine years of schooling or less) at the lowest
level, secondary education (high school or similar) for the middle group
and to academic degree at the highest level.

@@

In this section we analyse differences according to education level for the
period 1990--2013 for the whole population 40 years and older. Education is
a good indicator of both human and cultural capital, but aer also strongly
associated with income and social class. We present analyses on all cause
mortality (survival curves) as well as on the cause of death groups of
cardiovascular diseases and cancers (cumulative hazards). The pattern is
clear in total mortality as well as cause specific mortality, and this is
valid for both men and women, see
Figures~\ref{fig:ploteduc},~\ref{fig:ploteducheart}
and~\ref{fig:ploteduccancer}. Education had a strong, statistically
significant, effect on survival, as expected. Those with the lowest
education levels hade the highest mortality, while those with highest level
of education had by far the best survival. The differences must be
considered large, and it follows a continuous improvement according to
increasing levels.

```{r lissyoung3}
library(eha)
library(xtable)
liss4064 <- loTab[loTab$age < 64.5, ]
con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
##load("data/lisaold.rda")
load(con)
close(con)
```

```{r educ}
educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
educ$period <- factor(educ$period)
educ <- educ[!is.na(educ$educ), ]
levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
##educ$educ <- relevel(educ$educ, ref = "middle")
educ$civst <- relevel(educ$civst, ref = "married")
fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "male", ])
fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "female", ])
cols <- c("red", "darkgreen", "black")
```

```{r ploteduc, fig.cap="Survival by education and sex, 1990--2013."}
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
     col = cols, lwd = 1.5, lty = c(1, 2, 5), axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)

plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
     col = cols, lty = c(1, 2, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)

par(oldpar)
```

```{r tableeduc,results='asis'}
fit <- coxreg(Surv(enter, exit, event) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr, caption = "Effect of education on mortality in ages 40-90, 1990-2013.",
    label = "tab:tableeduc", digits = 4)
```

See Table ref{tab:tableeduc}.

```{r cardeduc}
ulcard <- unique(vb0$ULORSAK[vb0$heart])
educ$heart <- FALSE
educ$heart <- !is.na(educ$ULORSAK) & (educ$ULORSAK %in% ulcard) & educ$event
fitm <- coxph(Surv(enter, exit, heart) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "male", ])
fitf <- coxph(Surv(enter, exit, heart) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "female", ])
```

```{r ploteducheart, fig.cap="Cumulative hazards by education and sex, cardiovascular death, 1990--2013."}
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.9))
#axis(1)
#axis(2, at = c(0, 0.5, 1.0), las = 1)
#abline(h = 0.5, lty = 3)
#box()
legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
abline(h = 0)
plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
     col = cols, lty = c(1:2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.9))
#axis(1)
#axis(2, at = c(0, 0.5, 1.0), las = 1)
#abline(h = 0.5, lty = 3)
#box()
legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1:2, 5), lwd = 1.5)
abline(h = 0)
par(oldpar)
```

```{r tableeducheart,results='asis'}
fit <- coxreg(Surv(enter, exit, heart) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr, caption = "Effect of education on cardiovascular mortality in ages 40-90, 1990-2013.",
    label = "tab:tableeducheart", digits = 4)
```

See Table~\ref{tab:tableeducheart}.

```{r canceduc}
ulcard <- unique(vb0$ULORSAK[vb0$cancer])
educ$cancer <- FALSE
educ$cancer <- !is.na(educ$ULORSAK) & (educ$ULORSAK %in% ulcard) & educ$event
fitm <- coxph(Surv(enter, exit, cancer) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "male", ])
fitf <- coxph(Surv(enter, exit, cancer) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "female", ])
```

```{r ploteduccancer, fig.cap="Cumulative hazards by education and sex, cancer death, 1990--2013."}
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
#axis(1)
#axis(2, at = c(0, 0.5, 1.0), las = 1)
#abline(h = 0.5, lty = 3)
#box()
legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
abline(h = 0)
plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
     col = cols, lty = c(1, 2, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
#axis(1)
#axis(2, at = c(0, 0.5, 1.0), las = 1)
#abline(h = 0.5, lty = 3)
#box()
legend("topleft", legend = levels(educ$educ), col = cols, lty = c(1, 2, 5), lwd = 1.5)
abline(h = 0)
par(oldpar)
```

```{r tableeduccancer,results='asis'}
fit <- coxreg(Surv(enter, exit, cancer) ~ educ + civst + urban + sex, data = educ[!is.na(educ$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
ltx(fit, dr = dr, caption = "Effect of education on cancer mortality in ages 40-89, 1990-2013.",
    label = "tab:tableeduccancer", digits = 4)
```

See Table~\ref{tab:tableeduccancer}.

# The effect of Income, 1986--2005

```{r femeduc}
income <- lisaold[!is.na(lisaold$income), ]
income$period <- factor(income$period)
##income <- income[!is.na(income$income), ]
##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
##income$income <- relevel(income$income, ref = "low")
income$civst <- relevel(income$civst, ref = "married")
fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
             data = income[income$sex == "male", ])
fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
             data = income[income$sex == "female", ])
cols <- c("darkgreen", "red", "blue", "black")

```

```{r plotincome, fig.cap="Survival by income and sex, 1986--2005."}
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)

plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
axis(1)
axis(2, at = c(0, 0.5, 1.0), las = 1)
abline(h = 0.5, lty = 3)
box()
legend("bottomleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)

par(oldpar)
```

```{r tableinc}
fit <- coxreg(Surv(enter, exit, event) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
```

```{r tableincpr,results='asis'}
ltx(fit, dr = dr, caption = "Effect of income on mortality in ages 40-89, 1986-2005.",
    label = "tab:tableincpr", digits = 4)
```

See Table~\ref{tab:tableincpr}.

Finally, we present analyses of different income levels for the age group
40 and above and the period 1986--2005. Disposable individual income must
be considered as a robust measure of economic condition, and increases the
possibility to allocate these economic resources towards health promoting
living conditions and behaviours. We categorize the income according to
quartiles. As expected, we find that higher incomes result in lower
mortality, both when it comes to total mortality as to deaths from
cardiovascular diseases, see Figures~\ref{fig:plotincome}
and~\ref{fig:plotincheart}. This is also what could be expected, and
conforms to what we know of social health differences in present-day
Sweden. We also find indications of a similar difference for cancers but
less clear, particularly for male cancer mortality, see
Figure~\ref{fig:plotinccancer}. Notice however that we cannot establish a
causal relationship between income and mortality. Bad health may be a
confounder, causing both low incomes and higher death risks.

```{r heartincome}
ulcard <- unique(vb0$ULORSAK[vb0$heart])
income$heart <- FALSE
income$heart <- !is.na(income$ULORSAK) & (income$ULORSAK %in% ulcard) & income$event
fitm <- coxph(Surv(enter, exit, heart) ~ strata(income) + civst + urban,
             data = income[income$sex == "male", ])
fitf <- coxph(Surv(enter, exit, heart) ~ strata(income) + civst + urban,
             data = income[income$sex == "female", ])

```

```{r plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1986-2005."}
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 1.5))
#axis(1)
#axis(2, at = c(0, 0.5, 1.0), las = 1)
#abline(h = 0.5, lty = 3)
#box()
legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
abline(h = 0)
plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 1.5))
#axis(1)
#axis(2, at = c(0, 0.5, 1.0), las = 1)
#abline(h = 0.5, lty = 3)
#box()
legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
abline(h = 0)
par(oldpar)
```

```{r tableincheart}
fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
```

```{r tableincprheart,results='asis'}
ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.",
    label = "tab:tableincprheart", digits = 4)
```

See Table~\ref{tab:tableincprheart} for the effect of income on cardiovascular death.

```{r cancerincome}
ulcard <- unique(vb0$ULORSAK[vb0$cancer])
income$cancer <- FALSE
income$cancer <- !is.na(income$ULORSAK) & (income$ULORSAK %in% ulcard) & income$event
fitm <- coxph(Surv(enter, exit, cancer) ~ strata(income) + civst + urban,
             data = income[income$sex == "male", ])
fitf <- coxph(Surv(enter, exit, cancer) ~ strata(income) + civst + urban,
             data = income[income$sex == "female", ])

```

```{r plotinccancer, fig.cap="Cumulative hazards by income and sex, cancer death, 1990--2013."}
oldpar <- par(mfrow = c(1, 2))

plot(survfit(fitm, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Men",
     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
#axis(1)
#axis(2, at = c(0, 0.5, 1.0), las = 1)
#abline(h = 0.5, lty = 3)
#box()
legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
abline(h = 0)
plot(survfit(fitf, start.time = 40), fun = "cumhaz", mark.time = FALSE, main = "Women",
     col = cols, lty = c(1, 2, 4, 5), lwd = 1.5, xlab = "Age", ylab = "Cumulative hazards", ylim = c(0, 0.5))
#axis(1)
#axis(2, at = c(0, 0.5, 1.0), las = 1)
#abline(h = 0.5, lty = 3)
#box()
legend("topleft", legend = levels(income$income), col = cols, lty = c(1, 2, 4, 5), lwd = 1.5)
abline(h = 0)
par(oldpar)
```

```{r tableinccancer}
fit <- coxreg(Surv(enter, exit, heart) ~ income + civst + urban + sex, data = income[!is.na(income$civst), ],
              hazards = FALSE)
dr <- drop1(fit, test = "Chisq")
```

```{r tableincprcancer,results='asis'}
ltx(fit, dr = dr, caption = "Effect of income on cancer mortality in ages 40-89, 1986-2005.",
    label = "tab:tableincprcancer", digits = 4)
```

See Table~\ref{tab:tableincprcancer} for the income effect on cancer death.

>>>>

In the present paper, we take a long-term perspective on the development of
social inequalities in mortality in the adult and elderly population. The
areas under investigation are the Skellefteå and Umeå regions in the north
of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013
for Umeå. The main question we investigate is if inequality in mortality
between social classes increased among the adult and elderly population,
primarily benefiting the highest social positions and how this is connected
to the development of economic inequality in society. This issue is then
analysed according to gender, age-group (working age vs retired) and cause
of death (cardiovascular and cancer).

@@

@@
